---
title: "Intro to RMarkdown"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    highlight: tango
    toc: yes
    toc_depth: 5
    toc_float: yes
--- 

### Example heading 1
#### Example subheading 1

This is where you can write text with simple formatting:

1. **Example of bolded ordered list:** This text is no longer bolded.
2. *Now italicized:* No longer italicized.

* An unordered bullet point.
* Another one

Alright, it's time to embed some R code... 

`r paste("The time is", Sys.time())`

```{r}
# make a code chunk with three back ticks followed by an 'r' in braces {}
x <- rnorm(n = 100, mean = 10, sd = 5)
summary(x)
hist(x)
```

### Example heading 2
#### Example subheading 2

We can also read in the same simulated dataset, run the same code and embed Table 1 to share with collaborators.
```{r}
df <- readRDS(file.path("data", "simdata.rds"))
# load package (already installed) in order to use CreateTableOne function
library(tableone)
CreateTableOne(data = df,
               vars = c("age", "sex", "health_seek", "comorbid", "dx", "died"),
               strata = "vax",
               factorVars = c("health_seek", "comorbid", "dx", "died"),
               test = FALSE)
```
